### 2010-10-05 Thesis/Abstract.tex
- Update 5 October -
Added conclusion, abstract, resume and beginning of introduction



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@181 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\cleardoublepage

\begin{vcenterpage}
\noindent\rule[2pt]{\textwidth}{0.5pt}
\begin{center}
{\large\textbf{Real-time soft tissue modelling on GPU for medical simulation\\}}
\end{center}
{\large\textbf{Abstract:}}
Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla. Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.
\\
Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla. \\

Blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla. \\

Blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla.
\\
{\large\textbf{Keywords:}}
Medical simulation, soft tissue modelling, finite element method, GPU
\\
\noindent\rule[2pt]{\textwidth}{0.5pt}
\end{vcenterpage}

>>>>

\chapter*{Abstract}

\begin{center}
\large\textbf{Real-time soft tissue modelling on GPU for medical simulation\\}
\end{center}

\noindent \large\textbf{Abstract:}
Modelling the deformation of anatomical structures in real-time is a crucial problem in medical simulation. Because their shape and their constitution greatly differ, a unique model cannot deal with the variousness of the mechanical behaviours. Hence, we identified two major types of structures in human body, they can be either solid (brain, liver, prostate etc.) or hollow (colon, blood vessels, stomach etc.).

Our answer to this problematic is twofold. Our first contribution is an efficient GPU implementation of a non-linear, anisotropic and viscoelastic finite element model for solid organs. Our second contribution is a framework for real-time modelling of thin anatomical structures via a parallelisable co-rotational shell finite element formulation and a method to mesh a complex surface with curved shell elements.

Although the two soft tissue models are based on continuum mechanics for greater accuracy, they may both be employed to simulate the deformation of entire organs in real-time. Finally, their implementation into the open source framework SOFA will provide worldwide researchers with new models to assist in enhancing the realism of medical simulators.

\bigskip

\large\textbf{Keywords:}
Medical simulation, soft tissue modelling, finite element method, GPU, TLED, shell, co-rotational méthod

### 2010-10-05 Thesis/Chapter1.tex
- Update 5 October -
Added conclusion, abstract, resume and beginning of introduction



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@181 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\chapter{Introduction}
\label{chap:intro}
\begin{shortAbstract}
A short abstract for the upcoming chapter
\end{shortAbstract}

\section{Illustration Example}

>>>>

\chapter{Introduction to medical simulation}
\label{chap1}

\section{General context}

Virtually everyone knows this situation where a close relative has to go to the hospital, either for a simple routine check or a serious surgery. Hospitals are meant to be reinsuring, a refuge where people's health improves, where people get fixed. Obviously, patient safety is a critical concern in the medical industry. Yet, patients have been known to suffer injuries or even death due to errors of judgement or a lack of care and training. As an example, a report published in 2000 by the Institute of Medicine in the United States \citep{Kohn00} described two studies carried out in the 1990s based on large samples of hospital admissions. They found that the proportion of hospital admissions experiencing an adverse event, defined as causes of injuries by medical management, were 2.9 and $3.7\,$\%, respectively . The proportion of these events attributable to errors (that is, preventable adverse events, in other words medical errors) was 58 and $53\,$\%, respectively. Even when using the lower estimate, deaths due to preventable adverse events in 1997 in the US ($44\,000$) exceeded the deaths attributable to motor vehicle accidents ($43\,458$), breast cancer ($42\,297$) or AIDS ($16\,516$). Because many errors go unreported, detailed statistics on medical errors are fairly scarce (hospitals usually do not attempt to emphasize errors). Nevertheless, the problem is real. And this is not surprising, elaborate skills have to be mastered in the medical field. Moreover, achieving a high degree of personal competence is sometimes not sufficient as various members of staff must learn how to work as a team. In fact, one of the recommendations of this report \citep{Kohn00} was to \emph{establish interdisciplinary team training programs for providers that incorporate proven methods of training, such as simulation}. However, dispensing an appropriate training in such a complex and hazardous environment is very challenging. Because healthcare is a high risk industry (like aviation or military), training in the real world is too costly and dangerous. Consequently, various approaches were applied to teaching and training of medical practicians over history of healthcare.

\bigskip

The most basic form of medical simulators are simple models of human anatomy. Hundreds of years ago, representations in clay and stone were already used to demonstrate clinical features of disease states and their effects on humans. Models have been found from many cultures and continents. Nowadays, similar passive models are still used to help students learn the anatomy. In constrast, active models that attempt to reproduce living anatomy or physiology are recent developments. The first example was created in the early 1960s by Asmund Laerdal for training in mouth to mouth ventilation. This simulator of a dying victim not breathing and lacking a heart beat, became known as called Resusci-Anne and had been widely used for CPR training thanks to an internal spring attached to the chest wall \citep{Cooper04}. In the mid-1960s, Sim One became the first mannequin controlled by a computer. The chest was moved with breathing, the eyes blinked and the pupils could dilate for instance. But the computer technology was too expensive for commercialisation at the time and only one mannequin was built. Over time mannequins have obviously substantially improved and today such systems are essentially integrated into training centers that aim at recreating the operating room environment.

%- animal and fixed models used to reproduce some of the training objectives (but imperfect substitute and expensive)
%- animals, simulated patients, scenarios, computed-based

\section{Challenges for computer-based simulators}
%- Ultimate goal: needs to integrate patient-specific + physiology + pathologies for medical training, patient-specific planning and per-operative guidance		
%- Demands photorealistic rendering, accurate haptic feedback and (which is of prime interest to us): physically realist organ modelling in real-time.
%- Evolution: geometrical models => mass-springs => FEM => more elaborate constitutive laws / different type of structures
%- Contributions + plan

### 2010-10-05 Thesis/Chapter11.tex
- Update 5 October -
Added conclusion, abstract, resume and beginning of introduction



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@181 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

The objective of this PhD was to propose soft tissue models for computing the deformation of anatomical structures in real-time. Because the shape and the internal structure of these structures greatly differ, a unique model cannot deal with the variousness of the mechanical behaviours. Hence, we identified two major types of structures in human body, they can be either solid (brain, liver, prostate etc.) or hollow (colon, blood vessels, stomach etc.).

@@

\subsubsection*{Solid structure modelling.}
We implemented the total Lagrangian explicit dynamics (TLED) algorithm on GPU using CUDA. This finite element formulation is fully non-linear: it makes use of a non-linear strain measure (which allows large deformations) and embodies a non-linear constitutive law between strain and stress as characterised by \cite{Fung93} for soft-tissues. While the first GPU implementation of the TLED was carried out by \cite{Taylor07b} in the Cg language, our re-implementation in CUDA led to a threefold increase in performance. Overall, we observe a maximum speed improvement for GPU solution over CPU solution of up $56.3 \times$. My collaboration with Zeike Taylor continued and we enhanced the formulation by adding anisotropy and viscoelasticity. To the best of our knowledge, this work constitutes the first GPU implementation of a non-linear, anisotropic and viscoelastic finite element procedure. This formulation was eventually implemented in the open source framework for medical simulation SOFA and will be released publically with the upcoming 1.0 release of SOFA. By providing a fast and accurate non-linear FEM with various constitutive laws for soft tissue modelling to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators.

@@

The TLED algorithm offers an accurate modelling for soft tissues and its GPU implementation using CUDA is very fast. Hence, real-time computations can be achieved without any trade-off on the accuracy of the deformation. One of the features allowing such an efficiency is the use of an explicit time integration scheme. If this permits significant computational advantage, the use of an explicit integration scheme imposes that the time step of the simulation remains below a critical time step. This critical time step depends on two kind of criterion: the size of the elements and the material?s properties (Young?s modulus and Poisson?s ratio). Therefore, meshing of anatomical structures to be simulated must be accomplished with great care to insure well-shaped elements. Even more restricting, the use of the TLED formulation is not suitable for modelling stiff objects in real-time. Indeed, increasing the Young's modulus translates into a lower critical time step, which necessitates an intensification of the simulation's frequency to enforce real-time. In some applications, the value of the Poisson?s ratio may be relaxed from 0.49 to 0.45 for instance. However, if it allows to double the critical time step, this relaxation introduces additional inaccuracies.

@@

	\subsection{Real-time constraint: use of GPU acceleration}

@@

>>>>

The objective of this PhD was to propose soft tissue models for computing the deformation of anatomical structures in real-time. Because the shape and the internal nature of these structures greatly differ, a unique model cannot deal with the variousness of the mechanical behaviours. Hence, we identified two major types of structures in human body, they can be either solid (brain, liver, prostate etc.) or hollow (colon, blood vessels, stomach etc.).

@@

\subsubsection*{Solid structure modelling.}
We implemented the total Lagrangian explicit dynamics (TLED) algorithm on GPU using CUDA. This finite element formulation is fully non-linear: it makes use of a non-linear strain measure (which allows large deformations) and embodies a non-linear constitutive law between strain and stress as characterised by \cite{Fung93} for soft-tissues. While the first GPU implementation of the TLED was carried out by \cite{Taylor07b} in the Cg language, our re-implementation in CUDA led to a threefold increase in performance. Overall, we observe a maximum speed improvement for GPU solution over CPU solution of up $56.3 \times$. My collaboration with Zeike Taylor continued and we enhanced the formulation by adding anisotropy and viscoelasticity. To the best of our knowledge, this work constitutes the first GPU implementation of a non-linear, anisotropic and viscoelastic finite element procedure. This formulation was eventually implemented in the open source framework for medical simulation SOFA and will be released publically with the upcoming 1.0 release of SOFA. 
%By providing a fast and accurate non-linear FEM with various constitutive laws for soft tissue modelling to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators.

@@

The TLED algorithm offers an accurate modelling for soft tissues and its GPU implementation using CUDA is very fast. Hence, real-time computations can be achieved without any trade-off on the accuracy of the deformation. One of the features allowing such an efficiency is the use of an explicit time integration scheme. If this permits significant computational advantage, the use of an explicit integration scheme imposes that the time step of the simulation remains below a critical time step. This critical time step depends on two kind of criterion: the size of the elements and the material?s properties (Young's modulus and Poisson's ratio). Therefore, meshing of anatomical structures to be simulated must be accomplished with great care to insure well-shaped elements. Even more restricting, the use of the TLED formulation is not suitable for modelling stiff objects in real-time. Indeed, increasing the Young's modulus translates into a lower critical time step, which necessitates an intensification of the simulation's frequency to enforce real-time. In some applications, the value of the Poisson's ratio may be relaxed from 0.49 to 0.45 for instance. However, if it allows to double the critical time step, this relaxation introduces additional inaccuracies.

@@

	\subsection{Use of GPU acceleration}

One of the strongest constraint in medical simulation is the demand for real-time computations. This constraint is what makes all tasks in a simulator so challenging (collision detection, contact and soft tissue modelling, haptic feedback, photorealistic rendering etc.). In fact, this is at the origin of the multitude of simplifications proposed in simulation for modelling the biomechanics of anatomical structures. The lack of accuracy induced by these simplifications is of great concern to us. In this PhD, a part ot the solution was to accelerate the computations by using GPU. While we have showed that this approach can work very well, we need to keep in mind that GPU acceleration is not the panacea. Firstly, a standard is yet to be imposed in GPGPU. While OpenCL is openly targeted as a multi-vendor and multi-platform API, its usage remains very scarce nevertheless. In fact, the first experiments carried out within the SOFA framework revealed that OpenCL is not as optimised as CUDA for instance. Secondly, GPU programming is not simple to master. Although we have seen many improvements in flexibility of programmation over the last 3 or 4 years, a strong understanding of the underlying GPU hardware is still mandatory to create an optimal code. As an example, a GPU offers different types of memory, all with distinct pros and cons (strict access patterns, latency, limited size etc.), and choosing the best type of memory for each variable may be rather challenging to say the least. But most importantly, GPU acceleration is only interesting if the code is highly parallelisable. The difficulty is to devise a model that combines (1) an excellent capability at predicting the deformation of anatomical structures and (2) great parallel properties. Ideally, the parallelisation of the algorithm must be thought as part of the design process rather than seen as two unconnected steps. Even when the code is greatly parallelisable, one could face limited GPU hardware possibilities (as with concurrent writes for instance).

GPU acceleration is all but a straightforward process. One must often accomplish a considerable amount of work to achieve a substantial gain. However, a decent GPU implementation may allow improvements of up to one or two orders of magnitude. For instance, our CUDA implementation of the TLED algorithm was about $ 56 \times $ faster than an equivalent CPU implementation on 2008 hardware.

@@

Earlier this year, NVIDIA released a new GPU architecture called \emph{Fermi}. Fermi features several major innovations for GPGPU computing like cache (faster memory access in some cases), unified address space (easier to run C and C++ programs), debugging support, faster atomic instructions, compliance with the newly revised IEEE 754 floating point standard, 3 times the amount of shared memory, etc. In theory, this series of improvement should boost the performance of the second and third TLED implementations (see section \ref{chap6:secondTLED}) and we should investigate and compare the efficiency of the three designs on this new hardware.

In a short term perspective, our co-rotational shell FEM algorithm could also be implemented on GPU. We already assessed in section \ref{chap9:GPU} that there was no major obstacle which would prevent a parallel implementation. Nevertheless, an optimal GPU implementation will demand careful design to take full advantage of the hardware.

Although the meshing of complex surfaces with curved shell elements was not the heart of this PhD, our naïve approach could be enhanced by focusing on element quality. One simple solution would be to add a relaxation phase where a repulsion force is applied between each vertex (constrained to remain on the surface of the object). This physical approach would allow a more even distribution of the vertices across the surface and would contribute to obtain better quality elements overall. While a more mathematical method (a derivative of advancing front techniques for instance) could probably be formulated, the problem is complex and could account for another PhD.

Finally, all the contributions of this PhD could contribute to enhance the complexity of anatomical structure modelling through combining solid and hollow structures. While we presented our findings on the coupling between solid (tetrahedra) and thin objects (shells) via the simulation of an angioplasty procedure, these results are still preliminary and much work remains to be done. In particular, the validity of the interaction and the accuracy of the overall deformation of the blood vessel were not quantitatively evaluated.

As a conclusion, the TLED algorithm and our co-rotational shell FEM are respectively the first non-linear FEM for solid structures and the first model featuring bending for hollow structures to be implemented into the open source framework SOFA. Although they are based on continuum mechanics, they may both be employed to simulate the deformation of entire organs in real-time. By providing efficient FEM algorithms for all types of anatomical structures to worldwide researchers, we thus hope to assist in enhancing the realism of medical simulators.

### 2010-10-05 Thesis/Resume.tex
- Update 5 October -
Added conclusion, abstract, resume and beginning of introduction



git-svn-id: svn+ssh://scm.gforge.inria.fr/svnroot/sofa-shells@181 7b2761c9-2758-49ef-8c29-9c9f1e1aad1b

\large\textbf{Simulation de tissus mous en temps réel sur GPU pour la simulation médicale} \\

Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla. Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.
\\
Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla. \\

Blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla. \\

Blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla, blablablablabla bla blabla bla blablablabla bla.Blablabla bla blabla blablablabla.

>>>>

\begin{center}
\large\textbf{Simulation en temps réel de tissus mous sur GPU pour la simulation médicale\\}
\end{center}

\noindent \large\textbf{Résumé:}
Modéliser la déformation de structures anatomiques en temps réel est un problème crucial en simulation médicale. En raison des grandes différences existantes dans leur forme et leur constitution, un modèle unique est insuffisant face à la variété des comportements mécaniques. Par conséquent, nous avons identifié deux principaux types de structures: les organes pleins (cerveau, foie, prostate etc.) et les organes creux (colon, vaisseaux sanguins, estomac etc.).

Notre réponse à cette problématique est double. Notre première contribution est une implémentation GPU d'un modèle éléments finis qui est non-linéaire, anisotropique et viscoélastique pour les structures pleines. Notre seconde contribution est un environnement pour modéliser en temps réel les structures fines via un modèle parallèlisable et co-rotationnel utilisant des éléments coques et une approche pour mailler une surface complexe avec des éléments coques courbes.

Bien que les deux modèles de tissus sont basés sur la mécanique continue pour une meilleure précision, ils sont tous les deux capables de simulation la déformation d'organes en temps réel. Enfin, leur implémentation dans l'environnement open source SOFA rendra publique ces deux modèles afin de participer à l'amélioration du réalisme des simulateurs médicaux.

\bigskip

\large\textbf{Mots-clés:}
Simulation médicale, modélisation de tissus mous, méthode des éléments finis, GPU, TLED, coques, méthode co-rotationnelle
